---
title: "Chloropleth Maps in R"
output: pdf_document
---

# Chloropleth Maps

We've probably all seen more than our fair share of COVID dashboards over the past 16+ months. With that said, some of the state-by-state and county-by-county visualizations can be pretty powerful. In today's coding club, we're going to dig into how to create these visualizations, called chloropleth maps, using a couple different data sets.


Much of today's session is based upon this tutorial:

https://urban-institute.medium.com/how-to-create-state-and-county-maps-easily-in-r-577d29300bb2

There are other ways to create chloropleth maps in R, but for simplicity's sake, we'll focus on their method.


First, let's load some libraries we'll use. We all know tidyverse, but we're also going to load a couple packages created by the Urban Institute.

```{r}


library(tidyverse)
library(urbnmapr)
library(urbnthemes)
set_urbn_defaults(style = "print")

```


The urbnmapr package has a couple datasets that contain spatial information on states and counties:

```{r}

states

counties
```


Now, let's take a look at a dataset showing the Oregon vaccination rate by county as of last week (source: Oregonian):

```{r}

orCvData<-read_csv("OR_county_vax_cases.csv")
orCvData

```

Now in order to apply this data to a map, we first have to merge the covid dataset to one of the spatial datasets provided by urbnmapr. To do this, we need to find a match point, where both datasets have identical column values. Typical match points would either by the county name, or the county FIPS (Federal Information Processing Standards), which is basically a numerical identifier for each county.

Since our covid dataset does not contain a column for FIPS, we need to use county name instead. A problem, though, is that our covid dataset lists counties as single words (e.g. "Multnomah"), while the counties dataset lists them as two (e.g. "Multnomah County"). So, we need to get creative and 'wrangle' one of the datasets so we get a perfect match. Below, we'll use the 'mutate' function to create a new column in the counties dataset that lists counties as a single word. While we're at it, we'll create a subset of the counties dataset so it just contains Oregon counties.

```{r}

# create a subset of just Oregon counties
oregonCounties<-filter(counties, state_name=="Oregon")

# create a new column "County" that lists couties as a single word
oregonCounties<-oregonCounties %>%mutate(County=(str_split(county_name, " ", simplify = TRUE)[,1]))

oregonCounties
```

Now we're ready to merge the data using the "left_join" function:


```{r}


fullDataOr<-left_join(x=orCvData, y=oregonCounties, by=c("County"="County"))

fullDataOr



```

Finally, we can now take this combined data, and use ggplot to generate a map:


```{r}

orMap <-fullDataOr %>% 
  ggplot(mapping = aes(long, lat, group = group, fill = orcv$`Percentage of 18+ vaccinated`)) +
  geom_polygon(color = "#ffffff", size = .25) +
  scale_fill_gradient(labels = scales::percent,
                       guide = guide_colorbar(title.position = "top")) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme(legend.title = element_text(),
        legend.key.width = unit(0.5, "in")) +
  labs(fill = "Percent of 18+ Vaccinated") 


orMap

```




Now, can you create a map showing vaccination rates by state for August 9th, 2021, using the us_state_vaccinations.csv file?

The strategy would be:


* Read the csv file into a variable, and limit by date (using filter)
* Look at the data columns, and the data in the 'states' data set. Is there a match? Get creative if necessary!
* Merge the data using left_join
* Generate the map


```{r}






```


